```

```{r eval=TRUE, echo=FALSE, message=FALSE, purl=FALSE}
{
   options(width = 90)
   Sys.setenv(TZ = "UTC")
   library(datadr)
   library(trelliscope)
   library(cyberTools)
   setwd("~/Documents/Code/vastChallenge")
   cl <- makeCluster(8)
   clc <- localDiskControl(cluster = cl)
   bbRaw <- ddf(localDiskConn("data/bbRaw"))
   bbByHost <- ddo(localDiskConn("data/bbByHost"))
   nfBbByHost <- ddo(localDiskConn("data/nfBbByHost"))
}
```

## Network Health Data ##

### BB Exploration ###

Load the data and look at a subset...

```{r nhDataLoad, eval=TRUE, echo=TRUE, message=FALSE}
bbRaw <- ddf(localDiskConn("data/bbRaw"))
head(bbRaw[[1]][[2]][,-5])
```

...

### BB By Host Division ###

It turns out this data does not contain actual IPs, but instead the `hostname`, which is available in `hostList` which we can use to map to an IP address.  

```{r hostFreq, eval=TRUE, echo=TRUE}
bbHostFreq <- summary(bbRaw)$hostname$freqTable
head(bbHostFreq)
```

```{r eval=FALSE, echo=FALSE, purl=FALSE}
{
   tmp <- merge(bbHostFreq, hostList, by.x = "value", by.y = "hostName", all.x = TRUE)
   subset(tmp, is.na(tmp$IP))
   tmp <- tmp[order(tmp$Freq),]
   xyplot(Freq ~ 1:nrow(tmp), data = tmp, groups = type, auto.key = TRUE)
}
```

So to divide the data by host in a way that is similar to our netflow division, we need to first merge the `hostList` to get the actual IP.

Also note that there is some redundancy in columns.  For a given record, for example, if `connMade` is specified, then several of the preceding columns are always `NA`.  We can basically break the grouping of reported fields down into the following list:

```{r fields, eval=TRUE, echo=TRUE}
fields <- list(
   list(name = "disk", fields = "diskUsagePercent"),
   list(name = "page", fields = "pageFileUsagePercent"),
   list(name = "proc", fields = c("numProcs", "loadAveragePercent", "physicalMemoryUsagePercent")),
   list(name = "connMade", fields = "connMade")
)
```

We also want to exclude the big IPs like we did for netflow.  

Here we apply a `preTransFn` to merge in `hostList`, and a `postTransFn` to be applied to the final groupings to get things into a per-field format for simpler structure and smaller data size.

```{r bbByHost, eval=FALSE, echo=TRUE}
bigIPs <- c("172.20.0.15", "172.20.0.4", "172.10.0.4", "172.30.0.4")

bbByHost <- divide(bbRaw, by = "hostIP",
   preTransFn = function(x) {
      x <- merge(x, hostList, by.x = "hostname", by.y = "hostName", all.x = TRUE)
      x <- subset(x, !x$IP %in% bigIPs)
      names(x)[names(x) == "IP"] <- "hostIP"
      x
   },
   postTransFn = function(x) {
      x <- x[order(x$time),]
      res <- list()
      for(fld in fields) {
         ind <- which(!is.na(x[[fld$fields[1]]]))
         if(length(ind) > 0) {
            nms <- c("receivedfrom", "time", fld$fields)
            res[[fld$name]] <- x[ind, nms]
         }
      }
      res
   },
   output = localDiskConn("data/bbByHost"),
   control = clc
)
bbByHost <- updateAttributes(bbByHost, control = clc)
```

Lots more to look at here...

Here's an example of accessing and plotting data for a subset:


```{r name, eval=TRUE, echo=TRUE, fig.width=8, fig.height=3.5}
d <- bbByHost[[11]][[2]]
xyplot(connMade ~ time, data = d$conn)
xyplot(numProcs ~ time, data = d$proc)
xyplot(physicalMemoryUsagePercent ~ time, data = d$proc)
xyplot(loadAveragePercent ~ time, data = d$proc)
xyplot(diskUsagePercent ~ time, data = d$disk)
xyplot(pageFileUsagePercent ~ time, data = d$page)
```

### Joining with NetFlow ###

```{r join, eval=FALSE, echo=TRUE}
nfBbByHost <- drJoin(bb = bbByHost, nf = nfByHost, 
   output = localDiskConn("data/nfBbByHost"), control = clc)
```

Now we have a distributed data object where each subset is a list of netflow and network health data for one host

```{r joinEx, eval=TRUE, echo=TRUE}
str(nfBbByHost[[1]][[2]])
```

Now we can use this info in per-host analyses and vis...


