```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
{
   options(width = 90)
   Sys.setenv(TZ = "UTC")
   library(datadr)
   library(trelliscope)
   library(cyberTools)
   setwd("~/Documents/Code/vastChallenge")
   cl <- makeCluster(8)
   clc <- localDiskControl(cluster = cl)
   ipsRaw <- ddf(localDiskConn("data/ipsRaw"))
   ipsByHost <- ddf(localDiskConn("data/ipsByHost"))
   load("data/artifacts/ipsRows.Rdata")
}
```

## IPS Data ##

### IPS Exploration ###

```{r ipsEx, eval=TRUE, echo=TRUE}
head(ipsRaw[[1]][[2]])
```

...

```{r ipsFlags, eval=TRUE, echo=TRUE}
summary(ipsRaw)$flags$freqTab
```

...

### IPS By Host Division ###

Note that the source and destination IPs often contain the pre-[NAT](http://en.wikipedia.org/wiki/Network_address_translation)-ed IP.  We need to account for this when splitting the data up by host.  To do this, we merge in the `hostList` using the `externalIP` variable.

```{r ipsByHost, eval=FALSE, echo=TRUE}
ipsByHost <- divide(ipsRaw, by = "hostIP",
   preTransFn = function(x) {
      names(x)[names(x) == "destIp"] <- "destIpNat"
      # need to un-NAT the destination IP
      x <- merge(x, hostList[,c("IP", "externalIP")], by.x = "destIpNat", by.y = "externalIP", all.x = TRUE)
      names(x)[ncol(x)] <- "destIp"
      ind <- is.na(x$destIp)
      x$destIp[ind] <- x$destIpNat[ind]
      x <- getHost(x, src = "srcIp", dest = "destIp")
      subset(x, !x$hostIP %in% bigIPs)
   },
   output = localDiskConn("data/ipsByHost"),
   control = clc
)
ipsByHost <- updateAttributes(ipsByHost, control = clc)
```

Let's look at the distribution of number of rows for each host:

```{r ipsRows, eval=FALSE, echo=TRUE}
ipsRows <- recombine(ipsByHost, apply = nrow, combine = combRbind(), control = clc)
save(ipsRows, file = "data/artifacts/ipsRows.Rdata")
```

What does the distribution look like?

```{r ipsRowsPlot, eval=TRUE, echo=TRUE, fig.width=7, fig.height=5}
plot(log10(sort(ipsRows$val)))
```

There are some very large ones:

```{r name, eval=TRUE, echo=TRUE}
ipsBig <- subset(ipsRows, val > 90000)
merge(ipsBig, hostList, by. = "hostIP", by.y = "IP")
```

...

We can join this in with the other data sets by host as well...
